---
title: "Final Project Proposal"
author: "Vincent Chau, Christie Wong, Jimmy Leng, Arsenii Vasilev"
date: "2026-02-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Statement and Motivation

- define problem & research questions
  How do outside factors delay international flights?
- explain why the problem is important and interesting
  The problem is important because international aviation contributes trillions of dollars to the GDP of many countires, creating economic growth. Furthermore, many political figures rely on flights to travel from place to place. Having an understanding in why flights may be delayed can help prevent timimg issues as well as potentially save lives. This topic is interesting because we are discovering potential reasons in what factors are the main factors of delays. If we were to pinpoint specific indicators, we could possibly create predictive models based off these factors to ensure timely flights and improve efficiency in the aviation industry. 
- outline specific goals/hypotheses to be addressed
  We hypothesize that a combination of popular vacation destinations along with seasonal holidays and extreme weather conditions are high contributors to flight delays. 
- key questions you plan to explore
  How does the weather of particular regions affect flight delays?
  How do holidays/popular vacation times affect flight delays?
  Do popular vacation locations mean an increase in flight delays?
  Do specific weather conditions and seasons delay flights more than others?
  Do flight delays occur due to their destination or the origin of the airport?
  
  

## Data Description

  The dataset we use to support our proposal is from the nycflights13 package, which contains information on every flight from New York City in 2013 (link:https://nycflights13.tidyverse.org/). The datasets we focus on are the weather and flight datasets. 
  
  The flights dataset contains valuable information on factors that contribute to flight delays, including departure and arrival delays, and destinations. These variables are important because they are required to determine delay duration, and we hypothesize that destination popularity can cause delays.

  The weather dataset contains valuable information on factors that contribute to flight delays, including temperature, visibility, humidity, and precipitation. We consider these variables essential because temperature can affect delays when the weather is too cold or too hot; visibility is also a significant factor; humidity may also affect flight performance; and precipitation can be analyzed alongside visibility or on its own to determine whether excessive rainfall prevents flights. We will also use the variables year, month, and day, as both datasets include them, to facilitate data organization.

  We plan to combine all the data variables we want into a single dataset. We can easily incorporate weather and flight data because both datasets include year, month, and day. We then match the weather variables to the assigned dates and examine the weather for each flight. Together, the flights and weather datasets provide both delay outcomes and relevant environmental factors, making them sufficient for analyzing how weather conditions influence flight delays.

  In the dataset, some variables we examine have missing values, which can limit our analysis. To ensure data cleanliness, we will remove rows that lack the required values. For example, if there is a temperature data point missing for one row, we won’t consider that row, as we need to consider all the factors we chose for our data. 



## Exploratory Data Analysis (Extra Credit)

```{r cars, echo = false}
# packages needed to run the dataset, remove the # if missing any of the packages
#install.packages("tidyverse")
#install.packages("nycflights13")
#install.packages("dplyr")
library(nycflights13)
library(dplyr)
```

```{r}
#summary for the flights dataset in nycflights13
summary(flights)
```
# There is a lot of missing data for some of the variables in dep_delay, arr_time, arr_delay, and air_time. It appears that the variables with missing data are most likely due to flight cancellations, and some information can't be collected when flights are not in operation. We must consider that flights are canceled and either determine why they are canceled or exclude them from our project.



```{r}
#summary for the weather dataset in nycflights13
summary(weather)
```
#There are also missing values in the weather dataset; most of the missing data comes from wind_gust and pressure. For wind gusts, the large number of missing values is likely because winds are mostly steady, and for pressure, the flights are above ground, so sea-level pressure can't be recorded. We can't rely on these variables because they contain many missing values. There are other variables with missing values, but the number is slim to none, likely due to an error that prevented their recording. 



## Methodology & Analysis Plan

  To answer our questions, we will generate various plots to visualize our data and findings. Based on these graphs, we can make analyses and see if they support or contradict our hypothesis of popular vacation destinations, seasonal holidays, and extreme weather conditions highly contributing to flight departure and arrival delays. Because our project is concerned with the issue of flight delays, we will be focusing more on positive values for departure and arrival delay in minutes from the flights dataset in nycflights13. 

  Multiple scatterplots will see if weather variables like humidity, temperature, precipitation, and visibility have strong associations with flight delays. We can put regression lines on these scatterplots to visualize trends. We will create heatmaps to show which regions or destinations display the most amount of delays and their weather patterns by month on a map. 

  Line graphs can be made for each month that captures trends in average departure and arrival delays for each day. Some points in the graphs can be marked as holidays like December 25 for Christmas. This will help us see how holidays and vacation times affect flight delays. Multiple line graphs can be merged onto one plot to compare all 12 months with each other. This can get messy, so we can solely focus on specific months like the summer months against the winter months. We can also use line graphs to determine if certain destinations have spikes or drops in flight delays based on days or months. If the data is grouped by month and day intervals for each season (June 20 to September 22 for summer), we can measure which seasons experienced the most flight delays on average.

  In addition, bar graphs can display which destinations had the most flights and least flights by month and in the entire year. Creating bar graphs for each month helps us judge if certain places are more visited depending on the season. For example, we could expect LAX to be popular in June and July due to the sunny beaches in California. We can also use bar graphs to display the amount of flights departing from the three airports (EWR, JFK, LGA) in NYC. JFK is mostly international flights while LGA is mostly domestic flights. We can guess that JFK would have more flights and possibly would experience more delays. Grouped bar plots comparing average delay times between the three airports across the months would help us test that hypothesis.

  As we work on this project, we plan to meet online or in-person as a group to discuss goals and tasks we want to accomplish by the end of the week. Every member will do their assigned parts by the expected deadlines and communicate if there are any conflicts or other issues. This is so we stay on track and have a manageable workload as everyone has other responsibilities to tend to. It saves us the stress of cramming and finishing things at the last minute. 



## Alternative Strategies & Backup Plans

- discuss potential pitfalls/challenges you might encounter and propose clear alternative strategies + back up plans
1. Incompleteness or inconsistency of data:
  Our dataset might be incomplete or have missing values, which could limit the scope of our analysis.
  Backup plan: We cna create proxy variables to approximate the factors of interest. For example, instead of using detailed wether conditions, we cna replace them with binary variable, that will indicate if wheather conditions were good (1) for flight departures or bad (0). On another hand we can try to replace all the missing values and do a thorough data cleaning. 
2. High correlation of individual factors:
  There might occur difficulties in isolating individual factors due to high correlation.
  Backup plan: We can use statistical controls and simplified models to reduce confounding effects. For example, we can compare holiday delays woth non-holiday delays, delays with good weather conditions and delays with bad weather conditions separately.
3. Extreme broadness of scope:
The scope of our current dataset might be too wide for us to make accurate analysis.
Backup plan: We can narrow down our analysis to a smaller set of factors or move the focus onto smaller range of time and etc.
- list 2-3 other ideas you have brainstormed, each with a couple bullet points of details
1. Yelp reviews analysis
* [Yelp dataset](https://www.yelp.com/dataset)
* Is there any correlation between bad reviews and area codes? Or how likely bad reviews? Or how often reviews include photos and are they mostly bad?
* Types of analysis: Spatial, classification, probability analysis.
2. Uber rides analysis
*[Uber pickups](https://www.kaggle.com/fivethirtyeight/uber-pickups-in-new-york-city)
* What is the average trip length? What are of NYC uses Uber more frequently? How long does average ride takes?
* Types of analysis: Temporal, spatial or trip efficiency analysis.





