---
title: "Group 5 (VCAJ) Final Project Report"
author: "Vincent Chau, Christie Wong, Jimmy Ieng, Arsenii Vasilev"
date: "2026-03-13"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Different Factors on Flight Performance

## Project Description

  When traveling through airlines, there are various reasons to why the flight may be delayed. Whether the weather conditions are inadequate, the day you are flying out is the busiest, the destination airport may be packed and inefficient, anything can be a reason of delay. In our analysis, we will try and analyze a variety of these factors to see if there are any predictors for flight delay. In today's age flights are crucial moving the world forward. The world's most influential powers rely on aviation to visit various places for important business. Important resources and materials being sent via airlines each day, war relief to life saving organs. Any delays in these materials may be the line between life and death for people. Finding a plausible predictor for flight delays may drastically improve the efficiency of flights for millions of flyers each day.
  
### Questions we will be answering:
  
- Q1: Do flight delays occur due to destination or region?
  H[0]: Yes, destination/region does affect flight delay.
  H[a]: No, destination/region does not affect flight delay.
  
- Q2: Do certain months have an effect on flight delays?
  H[0]: Yes, certain months do affect flight delay.
  H[a]: No, certain months do not affect flight delay.
  
- Q3: Do specific weather conditions and seasons delay flights more than others?
  H[0]: Yes, specific weather conditions do affect flight delay.
  H[a]: No, specific weather conditions do not affect flight delay.
  
- Q4: Do flight delays occur due to the origin of the airport?
  H[0]: Yes, origin of the airport does affect flight delay.
  H[a]: No, origin of the airport does not affect flight delay.
  
- Q5: Do flight delays occur due to specific holidays?
  H[0]: Yes, specific holidays do affect flight delay.
  H[a]: No, specific holidays do affect flight delay.


## Data Description

  For our dataset, we used the “flights” and “weather” datasets from the “nycflights13” library. We decided that the important variables we wanted to look further into were month, day, year, destination, arrival delay, departure delay, temperature, precipitation, humidity, visibility, origin, and time. We decided to integrate the two datasets by the origin and time. Then we retained only the variables we considered important, as listed earlier, and used them to answer our questions. When looking at the dataset, we found some missing values, so we decided to remove any rows with NAs in any column. This reduced our observations from around 336k to 325k. 
  While working on the project, some limitations were encountered. Our first limitation was that we only had data from 2013, which means our dataset is limited to one year and includes only flights departing from NYC, so we can only analyze flights in NYC rather than flights across the whole country. One last limitation is that important data may have been missing or variables omitted.

```{r include = FALSE}
#install.packages("nycflights13")
#install.packages("tidyverse")
#install.packages("lubridate")
#install.packages("boot")
#install.packages("gridExtra")
#install.packages("ggdendro")
#install.packages("ggrepel")
#install.packages("caret")
```

```{r include = FALSE}
library(caret)
library(nycflights13)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
library(boot)
library(gridExtra)
library(ggdendro)
library(ggrepel)
```

```{r include = FALSE}
#filtering flights with the important variables we chose
Filtered_flights <- flights[, c("year", "month", "day", "dep_delay", "arr_delay", "dest", "origin", "time_hour")]
Filtered_flights %>% count(year, month, day) %>% filter(n > 1)
summary(Filtered_flights) # view(Filtered_flights)

#filtering weather with the important variables we chose
filtered_weather <- weather[,c(1,6,8,12,14,15)]
summary(filtered_weather) # view(filtered_weather)

#combining both filtered datasets by time_hour and origin
cleaned_weather_flights <- inner_join(Filtered_flights, filtered_weather, by = c("origin", "time_hour"))
cleaned_weather_flights <- drop_na(cleaned_weather_flights)
```


## Exploratory Data Analysis and Statistical Models

#### Q1: Do flight delays occur due to destination or region?

```{r echo = FALSE}
dest_delays <- cleaned_weather_flights |>
  group_by(dest) |>
  summarise(total_flights = n(),
            total_arr_delayed = sum(arr_delay > 15),
            total_dep_delayed = sum(dep_delay > 15),
            percent_arr_delay = total_arr_delayed/total_flights,
            percent_dep_delay = total_dep_delayed/total_flights)

# top 5 destinations with the most flights in 2013
busiest_dest_delay <- dest_delays |>
  arrange(desc(total_flights)) |>
  slice(1:5)

g1 <- ggplot(busiest_dest_delay, mapping = aes(x = reorder(dest, -percent_arr_delay))) +
  geom_col(aes(y = percent_arr_delay, fill = dest), position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Destination", 
       y = "% of Flights Delayed",
       title = "Arrival Delays in Top 5 Destinations") 

g2 <- ggplot(busiest_dest_delay, mapping = aes(x = reorder(dest, -percent_dep_delay))) +
  geom_col(aes(y = percent_dep_delay, fill = dest), position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Destination", 
       y = "% of Flights Delayed",
       title = "Departure Delays in Top 5 Destinations")

grid.arrange(g1, g2, ncol = 2)
```
 
  We hypothesized that airport destination would have an effect on flight delays, and we thought popular cities for vacation like Los Angeles may experience more delays due to the amount of planes flying in and out. We defined a popular or busy destination by their total number of flights. Based on the table above, there were many flights going into the airports: ATL, ORD, LAX, BOS, and MCO in 2013. This seems reasonable as they belong to states that many people like traveling or visiting to like Georgia, California, and Florida. 
  The bar graph visualizes the percent of flights delayed for arrivals and departures from the top 5 busiest destinations which are the airports: ATL, ORD, LAX, MCO, and BOS. Flights were considered late if they had over 15 minutes of delay. Despite having many flights in 2013, these locations performed relatively well, with only about 17% to 25% of their flights experiencing arrival delays and 17% to 23% of flights experiencing departure delays.

```{r echo = FALSE}
# Finding destinations with the most % of flights delayed (departure + arrival)
top5_arr_delayed <- dest_delays |>
  arrange(desc(percent_arr_delay)) |>
  slice(1:5)

top5_dep_delayed <- dest_delays |>
  arrange(desc(percent_dep_delay)) |>
  slice(1:5)

b1 <- ggplot(top5_arr_delayed, mapping = aes(x = reorder(dest, -percent_arr_delay))) +
  geom_col(aes(y = percent_arr_delay), fill = "salmon", position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Destination", 
       y = "% of Flights Delayed",
       title = "Top 5 Destinations w/ Arrival Delays") +
  theme(legend.position = "none")

b2 <- ggplot(top5_dep_delayed, mapping = aes(x = reorder(dest, -percent_dep_delay))) +
  geom_col(aes(y = percent_dep_delay), fill = "salmon", position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Destination",
       y = "% of Flights Delayed",
       title = "Top 5 Destinations w/ Departure Delays") +
  theme(legend.position = "none")

grid.arrange(b1, b2, ncol = 2)
```
  
  We decided to look at how the top 5 busiest destinations compared to actual destinations with the most amount of their flights delayed. For arrival delays, CAE, OKC, TUL, JAC, and TYS had the highest percentages of their flights delayed, ranging from about 40% to 65%. For departure delays, JAC, SBN, CAE, TUL, and OKC had the highest percentages of their flights delayed, ranging from 40% to 52%. These airports were also not that busy or popular as they only had a few hundred flights the entirety of 2013 from NYC, yet a large proportion of their flights were delayed in comparison to the busiest airports. 
  
  
```{r echo = FALSE}
# west - wash, oregon, cali, nev, utah, alaska, hawaii, montana, wy, colorado
# Northeast - connect, maine, massa, new_ham, ny, penn, rhode, vermont
# southwest - ari, new_mex, texas, ok
# southeast - ala, ark, flo, ga, ky, la, maryland, nc, sc, tenn, tx, ok, west virginia, va
# midwest - ill, indi, iowa, michi, minn, missouri, neb, ohio, wisconsin

west = c("SEA", "PDX", "BUR", "LAX", "LGB", "OAK", "PSP", "SAN", "SFO", "SJC", "SMF", "SNA",
         "LAS", "SLC", "ANC", "HNL", "BZN", "JAC", "DEN", "EGE", "HDN", "MTJ")
southwest = c("PHX", "ABQ", "AUS", "DFW", "HOU", "IAH", "SAT", "OKC", "TUL")
midwest = c("MDW", "ORD", "IND", "SBN", "DSM", "DTW", "GRR", "TVC", "MSP", "MCI",
            "STL", "OMA", "CAK", "CLE", "CMH", "DAY", "MKE", "MSN")
northeast = c("BDL", "BGR", "PWM", "ACK", "BOS", "MVY", "MHT", "ALB", "BUF", "ROC", "SYR",
              "PHL", "PIT", "PVD", "BTV")
southeast = c("BMH", "XNA", "EYW", "FLL", "JAX", "MCO", "MIA", "PBI", "RSW", "SRQ", "TPA",
              "ATL", "SAV", "CVG", "LEX", "SDF", "MSY", "BWI", "AVL", "CLT", "GSO", "ILM",
              "RDU", "BNA", "MEM", "TYS", "CAE", "CHS", "GSP", "MYR", "CRW", "CHO", "DCA",
              "IAD", "ORF", "RIC")
other = c("BQN", "PSE", "SJU", "STT")

region_data <- cleaned_weather_flights |>
  mutate(region = case_when(dest %in% west ~ "West",
                            dest %in% southwest ~ "Southwest",
                            dest %in% midwest ~ "Midwest",
                            dest %in% northeast ~ "Northeast",
                            dest %in% southeast ~ "Southeast",
                            dest %in% other ~ "Other"))
# linear model
summary(lm(arr_delay ~ region, data = region_data))
summary(lm(dep_delay ~ region, data = region_data))

not_canceled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))

inc_region <- cleaned_weather_flights |>
  mutate(region = case_when(dest %in% west ~ "West",
                            dest %in% southwest ~ "Southwest",
                            dest %in% midwest ~ "Midwest",
                            dest %in% northeast ~ "Northeast",
                            dest %in% southeast ~ "Southeast",
                            dest %in% other ~ "Other"))

not_canceled <- inner_join(not_canceled, inc_region)

delay_dest <- not_canceled %>%
  group_by(dest, region) %>%
  summarise(avg_dep_delay = mean(dep_delay, na.rm = TRUE),
                                 avg_arr_delay = mean(arr_delay, na.rm = TRUE))
delay_dest_scaled <- scale(delay_dest[, c("avg_dep_delay", "avg_arr_delay")])
km_fitted <- kmeans(x = delay_dest_scaled, centers = 5)
delay_dest$cluster <- as.factor(km_fitted$cluster)
ggplot(delay_dest, aes(x = avg_dep_delay, y = avg_arr_delay, color = cluster)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = region), size = 5) +
  labs(title = "Destination Cluster Based Off Average Delay", x = "Average Departure Delay", y = "Average Arrival Delay")


```
  Here we created a K-means cluster based off destinations. Here we first calculated the average departure delay as the x-axis, and the average arrival delay on the y-axis. Then we applied a scaling factor onto the values. Now having out scaled delay values, we applied a k-means clustering of 5 centers and plotted it onto a scatter plot. Here we used 4 clusters to view the varying groups of locations. From the graph, we can follow to the top right and observe that there generally is higher delay on average in the southwest part of the USA. This serves as a preliminary generally correlation between destination region and delay times.
  Linear regression models were used to further analyze the relationship between destinations and flight delays. Because there are more than 100 unique destinations in the dataset, each airport was grouped into one of 6 geographic regions within U.S. territory being the Midwest, Northeast, Southeast, Southwest, West, and Other( includes Puerto Rico and the Virgin Islands) to simplify and generalize our analysis. Our results showed that all regions significantly differed from the Midwest, except for the Southeast region in arrival delays. Overall, destination region is statistically significant and affects delays in both models, but it is a weak predictor on its own as the R-squared values show that less than 1% of the variation in delays can be explained by destination region.


#### Q2: Do certain months have an effect on flight delays?

```{r echo = FALSE}
month_delays <- cleaned_weather_flights |>
  group_by(month) |>
  summarise(total = n(),
            total_arr_delayed = sum(arr_delay > 15),
            total_dep_delayed = sum(dep_delay > 15),
            percent_arr_delay = total_arr_delayed/total,
            percent_dep_delay = total_dep_delayed/total) |>
  arrange(month) |>
  mutate(abbre_month = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))
```

```{r echo = FALSE}
month_graph_data <- month_delays |>
  select(month, abbre_month, percent_arr_delay, percent_dep_delay) |>
  pivot_longer(cols = c(percent_arr_delay, percent_dep_delay),
               names_to = "delay_type",
               values_to = "percent") |>
  mutate(delay_type = recode(delay_type,
                             "percent_arr_delay" = "Arr Delay",
                             "percent_dep_delay" = "Dep Delay"))

ggplot(month_graph_data, aes(x = reorder(abbre_month, month), y = percent, fill = delay_type)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Month",
       y = "% of Flights Delayed",
       fill = "Delay Type",
       title = "% of Delayed Flights by Month")
```

  We hypothesized that months were likely to affect flight delays as vacation months like June and December are likely to experience more congestion than any other time of the year. Based on the bar graph, the flight delay rates for arrivals tended to be about the same or greater than the flight delay rates for departures between each month. December, July, and June were the top three months with the highest flight delay rates while September, October, and November had the lowest flight delay rates.

```{r echo = FALSE}
summary(lm(arr_delay ~ factor(month), data = cleaned_weather_flights))
summary(lm(dep_delay ~ factor(month), data = cleaned_weather_flights))
```

  Linear regression models were used to analyze the relationship between month and flights delays in arrivals and departures. For the model on arrival delays, February, March, and August did not have a significant difference from January. Overall, like with the models for destination region, month is a statistically significant variable that affects delays; however it can only explain 1-2% of the variation in delays due to its low r-squared values. Months is a relatively weak predictor on its own, and this can be due to how multiple factors along with months can influence delays at the same time. With months, there may be seasons, holidays, or weather conditions at play.

#### Q3: Do specific weather conditions and seasons delay flights more than others?

#---------------------------Seasons-----------------------------------
```{r, echo=FALSE}
#Checking the percentage of flights by Seasons
#grouping months into seasons
flights_season <- cleaned_weather_flights %>%
  mutate(season = case_when(
    month %in% c(12, 1, 2) ~ "Winter",
    month %in% c(3, 4, 5) ~ "Spring",
    month %in% c(6, 7, 8) ~ "Summer",
    month %in% c(9, 10, 11) ~ "Fall"
  ))

flights_season

Percent_delay_by_Seasons <- flights_season %>% 
  group_by(season) %>%
  summarise(total_flights = n(),
            total_arr_delayed = sum(arr_delay > 15),
            total_dep_delayed = sum(dep_delay > 15),
            percent_arr_delay = total_arr_delayed/total_flights,
            percent_dep_delay = total_dep_delayed/total_flights)
Percent_delay_by_Seasons
#Spring: 24.2% Arrival Delay and 22.4% Departure Delay
#Summer: 28.6% Arrival Delay and 26.6% Departure Delay
#Winter: 26.2% Arrival Delay and 22.2% Departure Delay
#Fall: 15.7% Arrival Delay and 14.7% Departure Delay
```

#---------------------------looking at the linear model-----------------------------------

```{r,echo=FALSE}
MLR_season_arr_delay <-  lm(arr_delay ~ season, flights_season)
summary(MLR_season_arr_delay)
plot(MLR_season_arr_delay)
```
Assumption checks:
* Linearity - red line is horizontal, meaning there us a linear relationship
* Normality - not perfectly distributed, but most is distributed on the line, implying there are extreme outliers
* Constant Variance - shows a horizontal line, meaning the data is homoscedastic
* Outliers - There are a few outliers as there are points outside Cook's distance.

Based on the on the linear regression model, the seasons are a major factor as all have a p-value < 0.05.
* Fall: avgs -1 min delays
* Spring: avgs 8 min delay
* Summer: avgs 14 min delay
* Winter: avgs 10 min delay
The fall season may be the best to fly at.

```{r,echo=FALSE}
MLR_season_dep_delay <-  lm(dep_delay ~ season, flights_season)
summary(MLR_season_dep_delay)
plot(MLR_season_dep_delay)
```

Assumption checks:
* Linearity - red line is horizontal, meaning there us a linear relationship
* Normality - not perfectly distributed, but most is distributed on the line, implying there are extreme outliers
* Constant Variance - shows a horizontal line, meaning the data is homoscedastic
* Outliers - There are a few outliers as there are points outside Cook's distance.

  When looking at the linear model of the rain & visibility on the average mines of arrival delays it seems like rain and visibility has an effect to flight delays as the p-value <0.05, however they may not be significant as they both only explain 2% of the data. Also let it be known that in the bin[0.9,1.2) has a p-value >0.05, meaning that this area is not statistically significant, meaning there may not be enough evidence to conclude the association with delays in that category. For visibility, for every 1-mile increase in visibility, delays decrease by 2 minutes. For humidity, it is statistically significant however its effects are small compared to the other factors.

#------------------------------------------Weather Conditions------------------------------------------

```{r,echo=FALSE}
#Checking percentage of flights by rain 
# groups precipitation by 0.3 inch of rain
flights_rain <- cleaned_weather_flights %>%
  mutate(
    precip_bin = cut(
      precip, 
      breaks = seq(0, max(precip, na.rm = TRUE) + 0.3, by = 0.3),
      include.lowest = TRUE,
      right = FALSE
    )
  )
```

```{r,echo=FALSE}
#finding percentage of delays per 0.3 inches of rain
Avg_delay_per_rain <- flights_rain %>% 
  group_by(precip_bin) %>% 
  summarise(total_flights = n(),
            total_arr_delayed = sum(arr_delay > 15),
            total_dep_delayed = sum(dep_delay > 15),
            percent_arr_delay = total_arr_delayed/total_flights,
            percent_dep_delay = total_dep_delayed/total_flights)

Avg_delay_per_rain

#0.0 - 0.3 = barely rain: 23.6% Arrival Delay and 21.3% Departure Delay
#0.3 - 0.6 = little rain: 60.5% Arrival Delay and 50.5% Departure Delay
#0.6 - 0.9 = regular rain: 73.9% Arrival Delay and 45.6% Departure Delay
#0.9 - 1.2 = heavy rain: 45% Arrival Delay and 50% Departure Delay
#1.2 above = very heavy rain: 66.6% Arrival Delay and 88.8% Departure Delay

#Note: it seems like although there is a relationship with higher rain and delays, in the heavy rain category it has less delays than the rest

Graph_1 <- ggplot(Avg_delay_per_rain, aes(x = precip_bin)) +
  geom_col(aes(y = percent_arr_delay, fill = "Arrival Delay"), position = position_dodge()) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Precipitation Bin (inches)", y = "Percent Delayed", fill = "Delay Type",
       title = "Flight Delays vs Precipitation") +
  coord_flip()

Graph_2 <- ggplot(Avg_delay_per_rain, aes(x = precip_bin)) +
  geom_col(aes(y = percent_dep_delay, fill = "Departure Delay"), position = position_dodge()) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Precipitation Bin (inches)", y = "Percent Delayed", fill = "Delay Type",
       title = "Flight Delays vs Precipitation") +
  coord_flip()

grid.arrange(Graph_1, Graph_2, ncol = 2)

monthly_flights <- not_canceled %>%
  group_by(month) %>%
  summarise(avg_dep_delay = mean(dep_delay),
            avg_arr_delay = mean(arr_delay),
            number_flights = n())

monthly_weather <- weather %>%
  group_by(month) %>%
  summarize(avg_temp = mean(temp, na.rm = T),
            avg_hum = mean(humid, na.rm = T),
            avg_wind = mean(wind_speed, na.rm = T),
            avg_precip = mean(precip, na.rm = T),
            avg_press = mean(pressure, na.rm = T),
            avg_vis = mean(visib, na.rm = T))

joined_weather_delay <- inner_join(monthly_weather, monthly_flights)

joined_weather_delay_scale <- scale(joined_weather_delay[, -1])

matrix <- dist(joined_weather_delay_scale)

cluster <- hclust(matrix, method = 'average')

ggdendrogram(data = cluster) +
  ggtitle("Average Linkage Clusters Between Average Weather To Average Delay")
```

0.0 - 0.3 = barely rain: 23.6% Arrival Delay and 21.3% Departure Delay
0.3 - 0.6 = little rain: 60.5% Arrival Delay and 50.5% Departure Delay
0.6 - 0.9 = regular rain: 73.9% Arrival Delay and 45.6% Departure Delay
0.9 - 1.2 = heavy rain: 45% Arrival Delay and 50% Departure Delay
1.2 above = very heavy rain: 66.6% Arrival Delay and 88.8% Departure Delay

Note: it seems like although there is a relationship with higher rain and delays, in the heavy rain category it has less delays than the rest. Furthermore, the dengrogram describes clusters between average delay, as well as average weather conditions per month. They are then clustered via average linkage. This shows relations between seasons, where summer months like June and July have similar delay times as well as weather patterns. 

```{r,echo=FALSE}
#By visibility

#finding percentage of delays per visibility
Avg_delay_per_vis <- cleaned_weather_flights %>% 
  group_by(visib) %>% 
  summarise(total_flights = n(),
            total_arr_delayed = sum(arr_delay > 15),
            total_dep_delayed = sum(dep_delay > 15),
            percent_arr_delay = total_arr_delayed/total_flights,
            percent_dep_delay = total_dep_delayed/total_flights)

Avg_delay_per_vis 

Graph1 <- ggplot(Avg_delay_per_vis , mapping = aes(x = visib)) +
  geom_col(aes(y = percent_arr_delay, fill = "Arrival Delay"), position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percent of Flights Delayed", fill = "Delay Type",
       title = "Flight Delays by Visibility") 

Graph2 <- ggplot(Avg_delay_per_vis , mapping = aes(x = visib)) +
  geom_col(aes(y = percent_dep_delay, fill = "Departure Delay"), position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percent of Flights Delayed", fill = "Delay Type",
       title = "Flight Delays by Visibility") 

grid.arrange(Graph1, Graph2, ncol = 2)
#Note: based on the graph there is a correlation to delays and visibility however at the very low visibility there are some variables that are not high, probably due to not enough data/outliers
```
Note: based on the graph there is a correlation to delays and visibility however at the very low visibility there are some variables that are not high, probably due to not enough data/outliers.

```{r,echo=FALSE}
#By Temp
flights_temp <- cleaned_weather_flights %>%
  mutate(temp_group = case_when(
    temp < 40            ~ "Low",
    temp >= 40 & temp < 70 ~ "Medium",
    temp >= 70           ~ "High",
  ))
flights_temp

Avg_delay_per_temp <- flights_temp %>% 
  group_by(temp_group) %>% 
  summarise(total_flights = n(),
            total_arr_delayed = sum(arr_delay > 15),
            total_dep_delayed = sum(dep_delay > 15),
            percent_arr_delay = total_arr_delayed/total_flights,
            percent_dep_delay = total_dep_delayed/total_flights)

Avg_delay_per_temp

#Based on the data, there might not be a major correlation to delays
```

Based on the data, there might not be a major correlation to delays

```{r,echo=FALSE}
seasons_avg_temp <- flights_season %>% group_by(season) %>% summarise(
  avg_temp = mean(temp)
)

seasons_avg_temp
```
  Temperature can be grouped with season, as the delays for season & temp are similar. Temperature is strongly correlated with seasons, so including both in the same model may create redundancy.

```{r,echo=FALSE}
#By humidity
flights_hum <- cleaned_weather_flights %>%
  mutate(humidity_group = case_when(
    humid < 30             ~ "Low",
    humid >= 30 & humid < 60 ~ "Medium",
    humid >= 60             ~ "High",
  ))
flights_hum

Avg_delay_per_humid <- flights_hum %>% 
  group_by(humidity_group) %>% 
  summarise(total_flights = n(),
            total_arr_delayed = sum(arr_delay > 15),
            total_dep_delayed = sum(dep_delay > 15),
            percent_arr_delay = total_arr_delayed/total_flights,
            percent_dep_delay = total_dep_delayed/total_flights)

Avg_delay_per_humid

#Based on the data, there also might be a low correlation with humidity and delays
```
Based on the data, there also might be a low correlation with humidity and delays

#------------------Linear Model------------------------------------

```{r, echo=FALSE}
weather_lm_arr <- lm(arr_delay ~ precip_bin + visib + humid ,flights_rain)

summary(weather_lm_arr)
plot(weather_lm_arr)
```

Assumption checks:
* Linearity - red line is a straight line, meaning there is a linear relationship
* Normality - not perfectly distributed, but most is distributed on the line, implying there are extreme outliers
* Constant Variance - shows a small upward trend; there is slight heteroscedasticity in the data
* Outliers - There are a few outliers, as indicated by points with Cook's distance values.

  When looking at the linear model of the rain & visibility on the average mines of arrival delays it seems like rain and visibility has an effect to flight delays as the p-value <0.05, however they may not be significant as they both only explain 2% of the data. Also let it be known that in the bin[0.9,1.2) has a p-value >0.05, meaning that this area is not statistically significant, meaning there may not be enough evidence to conclude the association with delays in that category. For visibility, it seems like for every mile visibility increase the delays decrease by 2 minutes. For humidity, it is statistically significant however its effects are small compared to the other factors

```{r,echo=FALSE}
weather_lm_dep <- lm(dep_delay ~ precip_bin + visib + humid ,flights_rain)

summary(weather_lm_dep)
plot(weather_lm_dep)
```

Assumption checks:
* Linearity - red line is a straight line, meaning there is a linear relationship
* Normality - not perfectly distributed, but most is distributed on the line, implying there are extreme outliers
* Constant Variance - shows a small upward trend; there is slight heteroscedasticity in the data
* Outliers - There are a few outliers, as indicated by points with Cook's distance values.

  When looking at the linear model of the rain & visibility on the average minutes of departure delays, it seems like rain and visibility have an effect on flight delays as the p-value <0.05, they may not be significant as they both only explain 1% of the data. Once again, the bin[0.9,1.2) has a p-value >0.05, meaning that this area is not statistically significant, meaning there may not be enough evidence to conclude the association with delays in that category. For visibility, it seems that for every mile visibility increases, the delays decrease by 1.8 minutes. For humidity, it is statistically significant; however, its effects are small compared to the other factors. The reason for the bin[0.9,1.2) has a p-value >0.05 is most likely due to the sample size being way too small, as the bin has fewer than 10 flights for both arrival and departure delay.

```{r,echo=FALSE}
Percent_delay_by_Seasons %>%
  pivot_longer(cols = c(percent_arr_delay, percent_dep_delay),
               names_to = "delay_type",
               values_to = "percent_delay") %>%
  ggplot(aes(x = season, y = percent_delay * 100, fill = delay_type)) +
  geom_col(position = "dodge") +
  labs(
    title = "Flight Delay Percentage by Season",
    x = "Season",
    y = "Proportion of Flights Delayed (>15 minutes)",
    fill = "Delay Type"
  ) +
  scale_fill_manual(values = c("percent_arr_delay" = "steelblue",
                               "percent_dep_delay" = "tomato"),
                    labels = c("Arrival Delay", "Departure Delay")) +
  theme_minimal()

scale_y_continuous(labels = scales::percent)
```

*Do specific weather conditions and seasons delay flights more than others?*
  Both seasons and weather have statistically significant effects on flight delays. Flights in the summer are way worse than flights in fall, which seems to be like the best time to fly. When looking at the weather conditions, we can see that the temperature strongly correlates with the season, and the rain and visibility are evident. We can see that the more rain there is, the longer the delays in minutes, and that for each mile increase in visibility, the delay decreases by around 2 minutes. Humidity is also significant, but its effect on delays is very small. So weather and seasons may be a cause of delays, but their effects are not very big as they only explain a small variation of the data, with their explanation accounting for about 1-2% of the data, meaning there may be way more factors correlating with these variables in determining delays.


#### Q4: Do flight delays occur due to the origin of the airport?

```{r, echo=FALSE}

#Checking the Average arrival delay of origin
Avg_Origin_delays <- cleaned_weather_flights %>% 
  group_by(origin) %>% 
  summarise(total_flights = n(),
            total_arr_delayed = sum(arr_delay > 15),
            total_dep_delayed = sum(dep_delay > 15),
            percent_arr_delay = total_arr_delayed/total_flights,
            percent_dep_delay = total_dep_delayed/total_flights)
  
Avg_Origin_delays
#The 3 airports in New York seems to run efficiently
#EWR: 25.5% Arrival Delay and 24.5% Departure Delay
#JFK: 22.9% Arrival Delay and 20% Departure Delay
#LGA: 22.3% Arrival Delay and 18.8% Departure Delay

#gets the percentage of delays for each destination
percent_delay <- cleaned_weather_flights |>
  group_by(dest) |>
  summarise(total_flights = n(),
            total_arr_delayed = sum(arr_delay > 15),
            total_dep_delayed = sum(dep_delay > 15),
            percent_arr_delay = total_arr_delayed/total_flights,
            percent_dep_delay = total_dep_delayed/total_flights)


#compared to the nation average
#using data from destinations to find national average
Nation_Average_delays <- percent_delay %>% 
  summarise(total_flights = n(),
             mean_avg_arr = mean(percent_arr_delay),
      mean_avg_dep = mean(percent_dep_delay))
    
 Nation_Average_delays   

#National average: 26.1% Arrival Delay and 25% Departure Delay   
 
``` 
The 3 airports in New York seem to run efficiently.
* EWR: 25.5% Arrival Delay and 24.5% Departure Delay
* JFK: 22.9% Arrival Delay and 20% Departure Delay
* LGA: 22.3% Arrival Delay and 18.8% Departure Delay  
National average: 26.1% Arrival Delay and 25% Departure Delay   
 
```{r, echo=FALSE}
origin_lm_dep <- lm(dep_delay ~ origin ,cleaned_weather_flights)

summary(origin_lm_dep)
plot(origin_lm_dep)
```

Assumption checks:
* Linearity - red line is horizontal, meaning there is a linear relationship
* Normality - not perfectly distributed, but most is distributed on the line, implying there are extreme outliers.
* Constant Variance - shows a horizontal line, meaning the data is homoscedastic
* Outliers - There are a few outliers, as indicated by points with Cook's distance values outside the typical range.

```{r, echo=FALSE}
origin_lm_arr <- lm(arr_delay ~ origin ,cleaned_weather_flights)

summary(origin_lm_arr)
plot(origin_lm_arr)
``` 

Assumption checks:
* Linearity - red line is horizontal, meaning there is a linear relationship
* Normality - not perfectly distributed, but most is distributed on the line, implying there are extreme outliers
* Constant Variance - shows a horizontal line, meaning the data is homoscedastic
* Outliers - There are a few outliers, as indicated by points with Cook's distance values.

**Do flight delays occur due to the origin of the airport?**
  For the origin, the linear model shows a relationship for each of the three origin flights in NYC, as they are all statistically significant with p-values less than 0.05. EWR leading to 9 minutes of delays, JFK and LFA arrive 3 minutes early. However, the origins only explain less that 1% of the delays, meaning they are not a major factor in explaining the data alone.


#### Q5: Are flight delays more likely during the holidays?
```{r, echo=FALSE}
# Defining holday dates for year 2013
holiday_dates <- cleaned_weather_flights %>%
  mutate(
    date = as.Date(time_hour),
    holiday_period = case_when(
      date %in% seq(as.Date("2013-01-01"), as.Date("2013-01-01"), by = "day") ~ "New Year's",
      date %in% seq(as.Date("2013-05-25"), as.Date("2013-05-27"), by = "day") ~ "Memorial Day",
      date %in% seq(as.Date("2013-07-04"), as.Date("2013-07-07"), by = "day") ~ "Independence Day",
      date %in% seq(as.Date("2013-11-28"), as.Date("2013-12-01"), by = "day") ~ "Thanksgiving",
      date %in% seq(as.Date("2013-12-24"), as.Date("2013-12-31"), by = "day") ~ "Christmas/NYE",
      TRUE ~ "Non-Holiday"
    )
  )

# Summarizing delays for holidays
holiday_delays <- holiday_dates %>%
  group_by(holiday_period) %>%
  summarise(
    total_flights = n(),
    total_arr_delayed = sum(arr_delay > 15),
    total_dep_delayed = sum(dep_delay > 15),
    percent_arr_delay = total_arr_delayed / total_flights,
    percent_dep_delay = total_dep_delayed / total_flights
  ) %>%
  arrange(desc(percent_dep_delay))

holiday_delays
```


```{r, echo = FALSE}
# Comparing percentage of delays for each holday and non-holidays on average
h1 <- ggplot(holiday_delays, aes(x = reorder(holiday_period, -percent_arr_delay))) +
  geom_col(aes(y = percent_arr_delay, fill = holiday_period == "Non-Holiday"), position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Period", y = "% of Flights Delayed",
       title = "Holiday vs Non-Holiday:\n Arrival Delays") +
  theme(legend.position = "none") +
  coord_flip()


h2 <- ggplot(holiday_delays, aes(x = reorder(holiday_period, -percent_dep_delay))) +
  geom_col(aes(y = percent_dep_delay, fill = holiday_period == "Non-Holiday"), position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Period", y = "% of Flights Delayed",
       title = "Holiday vs Non-Holiday:\n Departure Delays") +
  theme(legend.position = "none") +
  coord_flip()

grid.arrange(h1, h2, ncol = 2)
```
  
  New Year's day has higher probability of arrival delays and period from Christmas to New Year's Eve has higher probability of departure delays. Based on that we can say that delays do happen more often during the winter holiday season.
  
```{r, echo = FALSE}
holiday_long <- holiday_delays %>%
  select(holiday_period, percent_arr_delay, percent_dep_delay) %>%
  pivot_longer(cols = c(percent_arr_delay, percent_dep_delay),
               names_to = "delay_type",
               values_to = "percent") %>%
  mutate(delay_type = recode(delay_type,
                             "percent_arr_delay" = "Arrival Delay",
                             "percent_dep_delay" = "Departure Delay"))

# Combined plot
ggplot(holiday_long, aes(x = holiday_period, y = percent, fill = delay_type)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Period",
       y = "% of Flights Delayed",
       fill = "Delay Type",
       title = "Flight Delays: Holiday vs Non-Holiday") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10))
```

Here we can see that only on New Year's arrival delays happen significantly more often than departure delays.

```{r echo = FALSE}
holiday_model_data <- holiday_dates %>%
  mutate(
    holiday = ifelse(holiday_period == "Non-Holiday", 0, 1),
    holiday_period = factor(holiday_period,
                            levels = c("Non-Holiday", "New Year's",
                                       "Memorial Day", "Independence Day",
                                       "Thanksgiving", "Christmas/NYE"))
  ) %>%
  drop_na(dep_delay)

m1 <- lm(dep_delay ~ holiday, data = holiday_model_data)

# Main: specific holiday periods
m2 <- lm(dep_delay ~ holiday_period, data = holiday_model_data)

summary(m1)
summary(m2)
```

```{r echo = FALSE}
model_metrics <- function(model, label) {
  tibble(
    Model  = label,
    R2     = summary(model)$r.squared,
    Adj_R2 = summary(model)$adj.r.squared,
    RMSE   = sqrt(mean(residuals(model)^2)),
    AIC    = AIC(model)
  )
}

bind_rows(
  model_metrics(m1, "M1: Binary Holiday"),
  model_metrics(m2, "M2: Specific Holidays")
)
```
  Two linear regression models were built to see if holidays are a good predictor of the flight delays. M1 considered only holiday and non-holiday seasons and M2 took each holiday into account besides the non-holiday period. Based on extremely R^2 values and high RMSE, it is clear that holidays is not a good predictor of arrival or departure delays.


## Model Implementation & Evaluation

```{r include = FALSE}
set.seed(42)
region_data <- na.omit(region_data)
region_glm_arr <- glm(arr_delay ~ region, data = region_data)
cv_result_region <- cv.glm(region_data, region_glm_arr, K = 10)$delta
cv_result_region

region_glm_dep <- glm(dep_delay ~ region, data = region_data)
cv_region_dep <- cv.glm(region_data, region_glm_dep, K = 10)$delta
cv_region_dep

month_glm_arr <- glm(arr_delay ~ factor(month), data = cleaned_weather_flights)
cv_result_month <- cv.glm(cleaned_weather_flights, month_glm_arr, K = 10)$delta
cv_result_month

month_glm_dep <- glm(dep_delay ~ factor(month), data = cleaned_weather_flights)
cv_month_dep <- cv.glm(cleaned_weather_flights, month_glm_dep, K = 10)$delta
cv_month_dep

weather_glm_arr <- glm(arr_delay ~ precip_bin + visib + humid, data = flights_rain)
cv_result_weather <- cv.glm(flights_rain,  weather_glm_arr, K = 10)$delta
cv_result_weather

weather_glm_dep <- glm(dep_delay ~ precip_bin + visib + humid, data = flights_rain)
cv_weather_dep <- cv.glm(flights_rain,  weather_glm_dep, K = 10)$delta
cv_weather_dep

origin_glm_arr <- glm(arr_delay ~ origin, data = cleaned_weather_flights)
cv_result_origin <- cv.glm(cleaned_weather_flights, origin_glm_arr, K = 10)$delta
cv_result_origin

origin_glm_dep <- glm(dep_delay ~ origin, data = cleaned_weather_flights)
cv_origin_dep <- cv.glm(cleaned_weather_flights, origin_glm_dep, K = 10)$delta
cv_origin_dep

season_glm_arr <- glm(arr_delay ~ season, data = flights_season)
cv_result_season <- cv.glm(flights_season,  season_glm_arr, K = 10)$delta
cv_result_season

season_glm_dep <- glm(dep_delay ~ season, data = flights_season)
cv_season_dep <- cv.glm(flights_season,  season_glm_dep, K = 10)$delta
cv_season_dep
```
  Cross validation was conducted on all factors of flight delays for arrivals and departures to evaluate the predictive performance of their linear regression models. Based on all the cross validation sets, all the models returned high mean squared error values ranging from about 1,500 to 2,000 for both arrival and departure delays, meaning that our models were not good at predicting delays. The model with the best predictive performance for arrival delays was the weather conditions model. It has the lowest MSE of 1,943 or RMSE of 44.07, where on average, predictions are about 44 minutes away from the actual arrival delays. Origin had the worst predictive performance for arrival delays with an MSE of 1,992 or RMSE of 44.63. For departure delays, the models with the best predictive performance were months and weather conditions with the lowest MSEs of about 1,581 or RMSEs of 39.76, where predictions are about 40 minutes away from actual departure delays on average. The worst predictive model for departures was region with an MSE of 1,604 or RMSE of 40.05.
```{r echo = FALSE}
set.seed(42)
cv_ctrl <- trainControl(method = "cv", number = 5)

cv_m1 <- train(dep_delay ~ holiday,
               data = holiday_model_data, method = "lm", trControl = cv_ctrl)

cv_m2 <- train(dep_delay ~ holiday_period,
               data = holiday_model_data, method = "lm", trControl = cv_ctrl)

resamples(list(M1 = cv_m1, M2 = cv_m2)) %>% summary()
```
  
  For holidays, two linear regression models were compared using 5-fold cross-validation. The model using specific holiday periods performed slightly better than the binary holiday model, but both models had extremely low explanatory power. 


## Results & Discussion

## Question 1:
  Destination serves as a significant predictor coming for its extremely low P-scores. However they currently cannot be used a predictor as its R^2 values are extremely low. This means that although significant, the current results cannot describe the variability of the results. A possible way to increase the R^2 score is to simply train on more data than what we have currently.

## Question 2:
  Results from the linear regression model reveals that months can be significant predictors for flight delay. Although having significance, it has the same issue as destination where the R^2 values of each month are extremely low. Therefore they are unable to explain the variability of the model. This may be because months follow similarly to seasons, holidays, and general trends that come from the year. 
  
## Question 3: 
  Both seasons and weather weather conditions are shown to be significant predictors for flight delay. Specifically in the summer where it is hotter and generally people go on vacation with either family or friends. Where summer is show to increase average flight delay by 12 minutes per unit. Furthermore, as seasons are correlated with weather, we can draw that the heavier the rain results in higher flight delays. This can be attributed in the real world through rain causing delays on the ground floor. 

## Question 4:
  From the regression models applied onto the data, we found that the three origin airports in New York City all show as significant predictors for flight delay. Each having their corresponding amounts of delay in minutes. While having low P-scores, their R^2 are nearly 0, meaning that the model cannot explain its variability.

## Question 5:
  As a result of applying models onto specific holidays of the year, we cannot conclude that holidays are a significant predictor of flight delays due to its low R^2 as well as extremely high RMSE value. This combination means that we will fail to reject the null hypothesis.
  
## Conclusion
  In the finaly anaylsis, we were able to confirm that there were in fact various factors that can be used to predictor flight delay. These include, origin airport, destination, weather, and season. However, due to the size of our data and lack of data, we ran into an issue of low explainability in regards to the models variability. Therefore, we cannot use the current models results as an accurate predictor. Some ways to potentially increase our R^2 values include, predicting on more data, transforming our data to reveal any non-linear trends, as well as better suited algorithms and models.


## Limitations

  While trying to answer these questions, we have run into various limitations that prevent us from fully gaining an answer. Although the datasets provides data to a great extent, it still is not enough data as a whole. The dataset only pertains to NYC while there are the other 49 states unaccounted for. Furthermore, this limits us to only New York's airlines, preventing us to anaylze possible predictors for airlines across the world. Additionally, the dataset omits various crucial variables about flights. Flights with N/A values contain no descriptive reason to the cancellation. This exclusion of data can cause omission bias into the models. Ultimately, regardless of the amount of data we have on these flights, it is truly not enough data to create concrete and feasible conclusions. 


## Team Member Contributions

* Vincent Chau - Team leader, set up github, planned what to do, additional models, conclusions,limitations, worked on final rmd.
* Christie Wong - Worked on data cleaning/EDA, worked on modeling for questions one and two, worked on final rmd.
* Jimmy Ieng - Worked on data cleaning/EDA, worked on modeling for questions three and four worked on final rmd.
* Arsenii Vasilev - Worked on EDA and modeling for question five, worked on final rmd. 

**Link to Github Repo**: 
  https://github.com/le3mosn/STAT167-GroupProject.git


